---
title: "Senior Capstone Project"
author: "Aidan Stewart"
date: '2024-01-15'
output: pdf_document
editor_options: 
  chunk_output_type: inline
---

#Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

# Installing Packages

```{r}
library(dplyr)
library(ggplot2)
library(rpart.plot)
library(reshape2)
library(mltsp)
library(e1071)
```

# Importing and Combining the .csv files (Baseball Reference)

```{r}

file_list <- list.files(pattern = "\\.csv$")

standard_data <- data_frame()

for (file in file_list) {
  current_data <- read.csv(file)
  file_name <- tools::file_path_sans_ext(basename(file))

  current_data$player_name <- file_name
  standard_data <- bind_rows(standard_data, current_data)
}

folder_path <- "C:/Users/Aidan/OneDrive/Documents/Senior Capstone Project/Advanced"

file_list_ADV <- list.files(path = folder_path, pattern = "\\.csv$", full.names = TRUE)

advanced_data <- data.frame()

for (file in file_list_ADV) {
  current_data <- read.csv(file)
  file_name <- tools::file_path_sans_ext(basename(file))
  extracted_name <- sub(" ADV$", "", file_name)

  current_data$player_name <- extracted_name
  advanced_data <- bind_rows(advanced_data, current_data)
}

common_columns = c("player_name", "Year", "Age", "Tm", "Lg")

combined_data <- merge(standard_data, advanced_data, by = common_columns, all.x = TRUE)

```

# Subsetting the data

```{r}

combined_data_subset <- subset(combined_data, !(Year == "162 Game Avg."))


combined_data_subset <- combined_data_subset[!grepl(c("Yrs"), combined_data_subset$Year, ignore.case = TRUE), ]


combined_data_subset <- combined_data_subset[, !(names(combined_data_subset) %in% c("Pos", "Awards", "Year.additional", "Lg", "EV", "HardH."))]


percentage_vars <- c("HR.", "SO.", "BB.", "LD.", "GB.", "FB.", "Pull.", "Cent.", "Oppo.", "cWPA", "RS.", "SB.", "XBT.")

for(var in percentage_vars) {
  combined_data_subset[[var]] <- as.numeric(gsub("%", "", combined_data_subset[[var]])) / 100
}

```

# Coorelation Plot

```{r}

# Identify numeric variables
numeric_vars <- sapply(combined_data_subset, is.numeric)

# Subset the data to include only numeric variables
numeric_data <- combined_data_subset[, numeric_vars]

# Remove rows with missing values
numeric_data_no_na <- na.omit(numeric_data)

# Calculate the correlation matrix excluding missing values
cor_matrix <- cor(numeric_data_no_na)

# Find correlations with 'WPA'
cor_with_WPA <- cor_matrix["WPA", ]

# Convert cor_with_WPA to a data frame with headers
cor_table <- data.frame(Variable = names(cor_with_WPA), Correlation = as.numeric(cor_with_WPA))

# Order the table by Correlation in descending order
cor_table <- cor_table[order(-cor_table$Correlation), ]

# Print the sorted table
print(cor_table)

# Convert correlation matrix to long format, excluding missing values
cor_long <- melt(cor_matrix, na.rm = TRUE)

# Create a heatmap
ggplot(cor_long, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, limit = c(-1, 1), space = "Lab", name="Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_fixed()


```

# Entering Baseball Savant data

```{r}

new_data <- read.csv("C:/Users/Aidan/OneDrive/Documents/Senior Capstone Project/New/stats_all.csv")

defensive_data <- read.csv("C:/Users/Aidan/OneDrive/Documents/Senior Capstone Project/New/outs_above_average.csv")

common_columns = c("ï..last_name..first_name", "player_id", "year")

combined_new_data <- merge(new_data, defensive_data, na.omit = TRUE, by = common_columns, all.x = TRUE)

selected_columns <- combined_new_data %>%
  select(ï..last_name..first_name, player_id, year, player_age, ab, pa, hit, home_run, k_percent, bb_percent, on_base_plus_slg, xba, xwoba, xiso, exit_velocity_avg, outs_above_average, fielding_runs_prevented, actual_success_rate_formatted)

selected_data <- na.omit(selected_columns)

selected_data

# Set up the plotting area
par(mfrow = c(1, 2))

# Bar plot before removing outliers
barplot(table(selected_data$player_age), col = "skyblue",
        xlab = "Age", ylab = "Player Count")

# Removing outliers
selected_data_filtered <- selected_data %>%
  filter(!(player_age <= 20) & !(player_age >= 40))

# Bar plot after removing outliers
barplot(table(selected_data_filtered$player_age), col = "skyblue",
        xlab = "Age", ylab = "Player Count")

table(selected_data_filtered$player_age)

# Reset the plotting area to default
par(mfrow = c(1, 1))



# 2020 Data change

selected_columns <- c("ab", "pa", "hit", "home_run", "outs_above_average", "fielding_runs_prevented")

filtered_data <- selected_data_filtered[selected_data_filtered$year == 2020, ]

# Multiply by 162 and divide by 60 for selected columns
filtered_data[, selected_columns] <- floor(filtered_data[, selected_columns] * 162 / 60)

selected_data_filtered[selected_data_filtered$year == 2020, ] <- filtered_data

selected_data_filtered



# Removing all seasons with less than 200 plate appearances
selected_data_filtered <- selected_data_filtered[selected_data_filtered$pa >= 200, ]

selected_data_filtered
```

# EDA for Baseball Savant

```{r}

plot(selected_data$player_age, selected_data$on_base_plus_slg, main = "OPS by Age", xlab = "Age", ylab = "OPS")


plot(selected_data$player_age, selected_data$tb, main = "tb by Age", xlab = "Age", ylab = "tb")


averages <- selected_data %>%
  group_by(player_age) %>%
  summarise(avg_on_base_plus_slg = mean(on_base_plus_slg))

ggplot(averages, aes(x = player_age, y = avg_on_base_plus_slg)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  labs(title = "Average of OPS by Age", x = "Age", y = "Average OPS")


averages <- selected_data %>%
  group_by(player_age) %>%
  summarise(avg_on_base_plus_slg = mean(outs_above_average, na.rm = TRUE))

ggplot(averages, aes(x = player_age, y = avg_on_base_plus_slg)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  labs(title = "Average outs above average by Age", x = "Age", y = "Average OAA")


plot(selected_data_filtered$player_age, selected_data_filtered$on_base_plus_slg, main = "OPS by Age", xlab = "Age", ylab = "OPS")

plot(selected_data_filtered$player_age, selected_data_filtered$tb, main = "tb by Age", xlab = "Age", ylab = "tb")

averages <- selected_data_filtered %>%
  group_by(player_age) %>%
  summarise(avg_on_base_plus_slg = mean(on_base_plus_slg))

ggplot(averages, aes(x = player_age, y = avg_on_base_plus_slg)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  labs(title = "Average of OPS by Age", x = "Age", y = "Average OPS")


averages <- selected_data_filtered %>%
  group_by(player_age) %>%
  summarise(avg_on_base_plus_slg = mean(outs_above_average, na.rm = TRUE))

ggplot(averages, aes(x = player_age, y = avg_on_base_plus_slg)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  labs(title = "Average outs above average by Age", x = "Age", y = "Average OAA")
```


# Making the quantiles

```{r}

calculate_quantiles <- function(data, variable_name) {
  quantile_col <- paste0(variable_name, "_quantile")
  
  data <- data %>%
    group_by(player_age) %>%
    mutate(!!quantile_col := ifelse(
      is.na(get(variable_name)),
      NA_real_,  # Handle NA values
      ifelse(get(variable_name) < quantile(get(variable_name), 0.25, na.rm = TRUE), 0,
             ifelse(get(variable_name) <= quantile(get(variable_name), 0.75, na.rm = TRUE), 1, 2)
      )
    )) %>%
    ungroup()  # Ungroup to remove grouping

  return(data)
}

# List of variables to loop through
variables_to_iterate <- c("xiso", "outs_above_average", "on_base_plus_slg", "xba")

# Loop through variables and apply the function
for (variable in variables_to_iterate) {
  selected_data_filtered <- calculate_quantiles(selected_data_filtered, variable)
}

selected_data_filtered <- selected_data_filtered %>%
  mutate(total_quantile = rowSums(select(., ends_with("_quantile")), na.rm = TRUE))

# Filter instances where total_quantile is 8
instances_with_total_8 <- selected_data_filtered %>%
  filter(total_quantile == 8)

# Print the instances
print(instances_with_total_8)

```

# Plotting the data for the 4 variables

```{r}

par(mfrow=c(2, 2))

for (variable in variables_to_iterate) {

  median_data <- aggregate(selected_data_filtered[[variable]], by = list(selected_data_filtered$player_age), FUN = median, na.rm = TRUE)
  names(median_data) <- c("player_age", "median_value")  
  
  plot(ggplot(selected_data_filtered, aes_string(x = "player_age", y = variable)) +
          geom_point() +
          geom_line(data = median_data, aes(y = median_value), color = "blue", size = 1.5) + # Add smooth curve representing median
          labs(title = paste("Scatter Plot of Players for", variable),
               x = "Player Age",
               y = variable) +
          theme_minimal())
}



# Add median line and quartiles to the plots

```

# Making the composite variable

```{r}

# Adding weights to the variables, I decided to do .6 for offensive stats and then .4 for defensive.
# Stating that the offensive part of the game is 60% of being a good player and then defensive is 40%.

weights <- c(.2, .4, .2, .2)

# These are the variables that I ended up iterating with.
variables_to_iterate <- c("xiso", "outs_above_average", "on_base_plus_slg", "xba")

# Normalize the four variables.
normalized_1 <- (selected_data_filtered$xiso - mean(selected_data_filtered$xiso)) / sd(selected_data_filtered$xiso)
normalized_2 <- (selected_data_filtered$outs_above_average - mean(selected_data_filtered$outs_above_average)) / sd(selected_data_filtered$outs_above_average)
normalized_3 <- (selected_data_filtered$on_base_plus_slg - mean(selected_data_filtered$on_base_plus_slg)) / sd(selected_data_filtered$on_base_plus_slg)
normalized_4 <- (selected_data_filtered$xba - mean(selected_data_filtered$xba)) / sd(selected_data_filtered$xba)

# Add the weights to each normalized variable
composite_score <- (weights[1] * normalized_1) +
                   (weights[2] * normalized_2) +
                   (weights[3] * normalized_3) +
                   (weights[4] * normalized_4)

# Put the variable into the dataset for each person
selected_data_filtered$CompositeScore <- composite_score

```



```{r}



# Group data by player_id
selected_data_filtered <- selected_data_filtered %>%
  group_by(player_id) %>%
  mutate(average_score = mean(CompositeScore),  # Calculate the average composite score for each player
         average_minus_year = CompositeScore - average_score) %>%
  ungroup()

# View the updated dataset
print(selected_data_filtered)


averages_minus_year <- selected_data_filtered %>%
  group_by(player_age) %>%
  summarise(avg_weightCompositeScore = mean(average_minus_year, na.rm = TRUE))


ggplot(averages_minus_year, aes(x = player_age, y = avg_weightCompositeScore)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  labs(x = "Age", y = "Normalized Composite score")



avg_weight_quad_model <- lm(avg_weightCompositeScore ~ poly(player_age, 2), data = averages_minus_year)

summary(avg_weight_quad_model)


# Define a sequence of ages
age_weight_seq <- seq(from = min(averages_minus_year$player_age), to = max(averages_minus_year$player_age), by = 1)

# Predictions for the quadratic model
avg_weight_quad_predictions <- predict(avg_weight_quad_model, newdata = data.frame(player_age = age_weight_seq))


# Plot the actual averages and predicted values with a bar plot
ggplot() +
  geom_bar(data = averages_minus_year, aes(x = player_age, y = avg_weightCompositeScore), stat = "identity", fill = "skyblue", color = "black") +
  geom_line(data = data.frame(player_age = age_weight_seq, predicted_avg_quad = avg_weight_quad_predictions),
            aes(x = player_age, y = predicted_avg_quad), color = "red", size = 1, linetype = "dashed") +
  labs(x = "Age", y = "Normalized Composite Score")



  
```

# Finding a model on the averages

```{r}

averages <- selected_data_filtered %>%
  group_by(player_age) %>%
  summarise(avg_CompositeScore = mean(CompositeScore, na.rm = TRUE))

ggplot(averages, aes(x = player_age, y = avg_CompositeScore)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  labs(x = "Age", y = "Average Composite score")

avg_model <- lm(avg_CompositeScore ~ player_age, data = averages)

ggplot(averages, aes(x = player_age, y = avg_CompositeScore)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed", formula = y ~ x) +  # Adding model line
  labs(x = "Age", y = "Average Composite score")

summary(avg_model)

```

# Quadratic model

```{r}

# Fit a quadratic regression model using the average composite score by age
avg_quad_model <- lm(avg_CompositeScore ~ poly(player_age, 2), data = averages)

# Define a sequence of ages
age_seq <- seq(from = min(averages$player_age), to = max(averages$player_age), by = 1)

# Predictions for the quadratic model
avg_quad_predictions <- predict(avg_quad_model, newdata = data.frame(player_age = age_seq))


# Plot the actual averages and predicted values with a bar plot
ggplot() +
  geom_bar(data = averages, aes(x = player_age, y = avg_CompositeScore), stat = "identity", fill = "skyblue", color = "black") +
  geom_line(data = data.frame(player_age = age_seq, predicted_avg_quad = avg_quad_predictions),
            aes(x = player_age, y = predicted_avg_quad), color = "red", size = 1, linetype = "dashed") +
  labs(x = "Age", y = "Average Composite Score")


summary(avg_quad_model)


# Residual analysis
hist(residuals(avg_quad_model), main = "Histogram of Residuals", xlab = "Residuals")

plot(fitted(avg_quad_model), residuals(avg_quad_model), 
     main = "Residuals vs. Fitted Values", xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)

qqnorm(residuals(avg_quad_model), main = "Normal Q-Q Plot")
qqline(residuals(avg_quad_model), col = "red", lty = 2)

plot(averages$player_age, residuals(avg_quad_model),
     main = "Residuals vs. Player Age", xlab = "Player Age", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)



```

# Cubic model

```{r}

# Fit a cubic regression model using the average composite score by age
avg_cubic_model <- lm(avg_CompositeScore ~ poly(player_age, 3), data = averages)

# Predictions for the cubic model
avg_cubic_predictions <- predict(avg_cubic_model, newdata = data.frame(player_age = age_seq))

# Plot the actual averages and predicted values with a bar plot
ggplot() +
  geom_bar(data = averages, aes(x = player_age, y = avg_CompositeScore), stat = "identity", fill = "skyblue", color = "black") +
  geom_line(data = data.frame(player_age = age_seq, predicted_avg_cubic = avg_cubic_predictions),
            aes(x = player_age, y = predicted_avg_cubic), color = "red", size = 1, linetype = "dashed") +
  labs(x = "Age", y = "Average Composite Score")

summary(avg_cubic_model)


```

# Use for loop to create individual player time series

```{r}

# Making a simpler dataset with just player_id, player_age, and composite_score

composite_data <- select(selected_data_filtered, player_id, player_age, CompositeScore)


# Assuming your data frame is called df

# Get unique player IDs
unique_players <- unique(composite_data$player_id)

# Create an empty list to store time series
time_series_list <- list()

# Iterate over unique player IDs
for (player_id in unique_players) {
  # Subset data for each player
  player_data <- composite_data[composite_data$player_id == player_id, ]
  
  # Create time series for player's age
  player_time_series <- ts(player_data$CompositeScore, start = player_data$player_age[1])
  
  # Add time series to the list
  time_series_list[[as.character(player_id)]] <- player_time_series
}

# Now time_series_list contains time series for each unique player ID

# Showing the data for the player Jose Abreu
time_series_list$"547989"

```

# Create NARX model for each time series

```{r}

model <- narx(time_series_list$"592743", svm, p = 1, freq = 1)

fcst <- forecast(model, h = 5)

plot(fcst)

```


# Create t-test boxplots where you take players with 

```{r}


delta_t_test <- function(data, age_1, age_2, var) {
  # Find unique player_ids with both age_1 and age_2
  players_with_both_ages <- unique(data$player_id[data$player_age == age_1 & data$player_id %in%  unique(data$player_id[data$player_age == age_2])])
  
  # Filter the data for these players
  filtered_data <- data[data$player_id %in% players_with_both_ages, ]
  
  # Create a new dataset with columns age_1_var, age_2_var, and delta
  age_1_var <- filtered_data[filtered_data$player_age == age_1, c("player_id", var)]
  colnames(age_1_var)[2] <- paste0("age_", age_1, "_", var)
  
  age_2_var <- filtered_data[filtered_data$player_age == age_2, c("player_id", var)]
  colnames(age_2_var)[2] <- paste0("age_", age_2, "_", var)
  
  new_data <- merge(age_1_var, age_2_var, by = "player_id")
  
  print(new_data)
  
  new_data$delta <- new_data[[paste0("age_", age_2, "_", var)]] - new_data[[paste0("age_", age_1, "_", var)]]
  
  # Check the number of observations
  cat("Number of observations:", nrow(new_data), "\n")
  
  # Perform t-test
  result <- t.test(new_data$delta)
  
  return(result)
}

# Example usage:
# Assuming your data is stored in a data frame called 'data'
# where 'player_id', 'age', and 'var' are column names
result <- delta_t_test(composite_data, 34, 37, "CompositeScore")

print(result)

# Interpret the results
cat("T-statistic:", result$statistic, "\n")
cat("Degrees of freedom:", result$parameter, "\n")
cat("P-value:", result$p.value, "\n")
cat("95% Confidence Interval for the Difference:", result$conf.int, "\n")
cat("Estimated Mean Difference (delta):", result$estimate, "\n")

```


# T-test loop

```{r}

# Define a function to run delta_t_test in a loop
run_delta_t_test_loop <- function(data, start_age, end_age, var) {
  results <- list()
  
  for (age_1 in start_age:end_age) {
    age_2 <- age_1 + 3
    
    cat("Running delta_t_test for ages:", age_1, "and", age_2, "\n")
    
    result <- delta_t_test(data, age_1, age_2, var)
    
    results[[paste0("Age_", age_1, "_", age_2)]] <- result
  }
  
  return(results)
}

# Call the function to run delta_t_test in a loop
results <- run_delta_t_test_loop(composite_data, 21, 36, "CompositeScore")

results

# Extract t-values from results
t_values <- sapply(results, function(x) x$statistic)

# Create a data frame from the t-values
t_values_df <- data.frame(Age_Pair = names(t_values), t_value = t_values)

# Print the data frame
print(t_values_df)


# Create plot
plot(t_values)


# Extract p-values from results
p_values <- sapply(results, function(x) x$p.value)

# Create a data frame from the t-values
p_values_df <- data.frame(Age_Pair = names(p_values), p_value = p_values)

# Print the data frame
print(p_values_df)


# Create plot
plot(p_values)



# Risk signing 26 year old


```



```{r}

# Convert t_values and p_values to data frames with appropriate index values
t_values_df <- data.frame(player_age = seq(21, 36), t_value = t_values)
p_values_df <- data.frame(player_age = seq(21, 36), p_value = p_values)

# Create plot for t_values
t_plot <- ggplot(t_values_df, aes(x = player_age, y = t_value)) +
           geom_point() +
           labs(x = "Player Age", y = "t-value")  # Rename the axis labels

# Create plot for p_values
p_plot <- ggplot(p_values_df, aes(x = player_age, y = p_value)) +
           geom_point() +
           labs(x = "Player Age", y = "p-value") +  # Rename the axis labels
           geom_hline(yintercept = 0.05, linetype = "solid", color = "black")  # Add a solid line at y = 0.05

# Display the plots
print(t_plot)
print(p_plot)


```